---
title: "p8105_hw2_ebm2142"
author: "Elise Mantell"
date: 2019-10-04
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(ggplot2)
```
### Problem 1

#### Read and clean the Mr. Trash Wheel sheet

```{r reading trash data}
Mr_TW = read_excel("./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx", range = "A2:N408", sheet = 1) %>% 
janitor::clean_names() %>% 
filter(dumpster != "NA") %>% 
mutate(sports_balls = as.integer(round(sports_balls))) %>% 
rename("weight" = weight_tons, "volume" = volume_cubic_yards)
```
#### Read and clean precipitation data for 2017 and 2018. For each, omit rows without precipitation data and add a variable year


```{r reading precip data}
#2017 data
precip_17 = read_excel("./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx", sheet = "2017 Precipitation", range = "A2:B14") %>% 
  janitor::clean_names() %>% 
  filter(total != "NA") %>% 
#adding variable
  mutate(year = 2017)

#2018 data
precip_18 = read_excel("./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",sheet = "2018 Precipitation", range = "A2:B14") %>% 
  janitor::clean_names() %>% 
  filter(total != "NA") %>% 
#adding variable
mutate(year = 2018)
```
#### Combine precipitation datasets and convert month to a character variable (the variable month.name is built into R and should be useful).

```{r combining and converting month variable}
precip_17_18 =
  full_join(precip_17, precip_18) %>% 
  mutate(month = month.name[month])
```

#### Write a paragraph about these data
* Be sure to note the number of observations in both resulting datasets
* Give examples of key variables
* What was the total precipitation in 2018? 
* What was the median number of sports balls in a dumpster in 2017? 

```{r describing}

# The number of observations in the precipitation dataset is
nrow(precip_17_18)

# The number of observations in the Mr Trashwheel dataset is
nrow(Mr_TW)

# The total precipitation in 2018 was
sum((pull(precip_18, total)))

# The median number of sports balls in a dumpster in 2017 was
median(pull(Mr_TW %>% filter(year == 2017), sports_balls))
```
### Problem 2

#### Clean the data in pols-month.csv. Use separate() to break up the variable mon into integer variables year, month, and day; replace month number with month name; create a president variable taking values gop and dem, and remove prez_dem and prez_gop; and remove the day variable.

```{r cleaning pols-month}
pols_month = read_csv("./data/pols-month.csv") %>%
  janitor::clean_names() %>% 
  separate("mon", into = c("year", "month", "day")) %>% 
  mutate(month = month.abb[as.integer(month)],month =  str_to_lower(month)) %>% 
  rename("gop" = prez_gop, "dem"= prez_dem)
```

#### Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.

```{r cleaning snp}
snp = read_csv("./data/snp.csv") %>% 
  janitor::clean_names() %>% 
  separate("date", into = c("day", "month", "year")) %>% 
  select("year", "month", "day", "close") %>% 
  mutate(month = month.abb[as.integer(month)],month =  str_to_lower(month))
```

#### Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.

```{r cleaning unemployment}
unemployment = read_csv("./data/unemployment.csv", col_types = "cdddddddddddd") %>% 
  janitor::clean_names() %>% 
  pivot_longer(jan:dec, 
  names_to = "month", values_to = "rate") 
```

#### Join the datasets by merging snp into pols, and merging unemployment into the result.

```{r merging snp into pols and unemployment into that}
snp_pols = left_join(snp, pols_month, by = c("year","month"))
snp_pols_unemp = left_join(snp_pols, unemployment, by = c("year","month"))
```
#### Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).

```{r describing combined dataset}
nrow(snp_pols_unemp)
range(pull(snp_pols_unemp, year))
```

##### The dataset “pols-month” contains 822 observations of 9 variables related to the number of national politicians who are democratic or republican at a given time ("gop" and "dem" variables indicating if the president was a republican or democrat, with similar variables for senate, house of representatives, and governors). The dataset "snp" contains 787 observations the position of the Standard & Poor’s stock market index (S&P) at close (represented by variable "close"). The dataset “unemployment” contains 68 observations of 13 variables, showing the unemployment percentage for each month of the year. The combined dataset containing all three of these includes 787 observations over the years 1950 to 2015. 

### Problem 3

#### Load and tidy the data. The names of a categorical predictor and the case structure of string variables changed over time; you’ll need to address this in your data cleaning. Also, some rows seem duplicated, and these will need to be removed

```{r load and tidy baby name data}
baby_name = read_csv("./data/Popular_Baby_Names.csv", col_types = "icccii") %>% 
janitor::clean_names() %>% 
distinct() %>% 
mutate_all(toupper) %>% 
rename("year" = year_of_birth, "name" = childs_first_name) %>% 
mutate(ethnicity = replace(ethnicity, ethnicity == "ASIAN AND PACI", "ASIAN AND PACIFIC ISLANDER")) %>% 
mutate(ethnicity = replace(ethnicity, ethnicity == "BLACK NON HISP","BLACK NON HISPANIC")) %>%
mutate(ethnicity = replace(ethnicity, ethnicity == "WHITE NON HISP", "WHITE NON HISPANIC"))
```

#### Produce a well-structured, reader-friendly table showing the rank in popularity of the name “Olivia” as a female baby name over time; this should have rows for ethnicities and columns for year. Produce a similar table showing the most popular name among male children over time.

```{r producing Olivia table}
baby_name %>%
  group_by(year, ethnicity, rank) %>% 
  filter(name == "OLIVIA") %>%
  knitr::kable(digits = 1)
```
```{r producing male top baby name table}
baby_name %>%
  group_by(year, ethnicity, name) %>% 
  filter(rank == 1, gender == "MALE") %>%
  knitr::kable(digits = 1)
```

#### Finally, for male, white non-hispanic children born in 2016, produce a scatter plot showing the number of children with a name (y axis) against the rank in popularity of that name (x axis)
