---
title: "p8105_hw2_ebm2142"
author: "Elise Mantell"
date: 2019-10-04
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(ggplot2)
```
### Problem 1

#### Read and clean the Mr. Trash Wheel sheet

```{r reading trash data}
Mr_TW = read_excel("./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx", range = "A2:N408", sheet = 1) %>% 
janitor::clean_names() %>% 
filter(dumpster != "NA") %>% 
mutate(sports_balls = as.integer(round(sports_balls))) %>% 
rename("weight" = weight_tons, "volume" = volume_cubic_yards)
```
#### Read and clean precipitation data for 2017 and 2018. For each, omit rows without precipitation data and add a variable year


```{r reading precip data}
#2017 data
precip_17 = read_excel("./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx", sheet = "2017 Precipitation", range = "A2:B14") %>% 
  janitor::clean_names() %>% 
  filter(total != "NA") %>% 
#adding variable
  mutate(year = 2017)

#2018 data
precip_18 = read_excel("./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",sheet = "2018 Precipitation", range = "A2:B14") %>% 
  janitor::clean_names() %>% 
  filter(total != "NA") %>% 
#adding variable
mutate(year = 2018)
```
#### Combine precipitation datasets and convert month to a character variable (the variable month.name is built into R and should be useful).

```{r combining and converting month variable}
precip_17_18 =
  full_join(precip_17, precip_18) %>% 
  mutate(month = month.name[month])
```

#### Write a paragraph about these data
* Be sure to note the number of observations in both resulting datasets
* Give examples of key variables
* What was the total precipitation in 2018? 
* What was the median number of sports balls in a dumpster in 2017? 

```{r describing}

# The number of observations in the precipitation dataset
nrow(precip_17_18)

# The number of observations in the Mr Trashwheel dataset
nrow(Mr_TW)

# The total precipitation in 2018 was
sum((pull(precip_18, total)))

# The median number of sports balls in a dumpster in 2017 was
median(pull(Mr_TW %>% filter(year == 2017), sports_balls))
```
### Problem 2

#### Clean the data in pols-month.csv. Use separate() to break up the variable mon into integer variables year, month, and day; replace month number with month name; create a president variable taking values gop and dem, and remove prez_dem and prez_gop; and remove the day variable.

```{r cleaning pols-month}
pols_month = read_csv("./data/pols-month.csv") %>%
  janitor::clean_names() %>% 
  separate("mon", into = c("year", "month", "day")) %>% 
  mutate(month = month.abb[as.integer(month)],month =  str_to_lower(month)) %>% 
  rename("gop" = prez_gop, "dem"= prez_dem)
```

#### Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.

```{r cleaning snp}
snp = read_csv("./data/snp.csv") %>% 
  janitor::clean_names() %>% 
  separate("date", into = c("day", "month", "year")) %>% 
  select("year", "month", "day", "close") %>% 
  mutate(month = month.abb[as.integer(month)],month =  str_to_lower(month))
```

#### Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.

```{r cleaning unemployment}
unemployment = read_csv("./data/unemployment.csv") %>% 
  janitor::clean_names() %>% 
  pivot_longer(jan:dec, 
  names_to = "month", values_to = "rate") 
```

#### Join the datasets by merging snp into pols, and merging unemployment into the result.


